A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-27 sections/Metatheory.tex
revision of 4.5

\section{Metatheory}

@@

\subsection{Decidability}

\bruno{Tom, this is the section that I feel less helpful. Maybe
you can look at this one more closely?}

The decidability theorem is:

\begin{theorem}[Decidability]
Given \emph{wf }$\Gm$, it is decidable whether $\Gm\redto\nil$ or not.
\end{theorem}

We have proven this theorem by a lexicographic group of induction measures\\
$\langle |\Gm|_e, |\Gm|_\Leftrightarrow, |\Gm|_\forall, |\Gm|_{\al}, |\Gm|_\to + |\Gm| \rangle$
on the worklist $\Gm$.
$|\cdot|_e$ and $|\cdot|_\Leftrightarrow$ measure the total ``size'' of terms
and the total ``difficulty'' of judgments, respectively.
$|\cdot|_\forall$ and $|\cdot|_\to$ counts the total number of
universal quantifiers and function types, respectively. Finally,
$|\cdot|_{\al}$ counts the number of existential variables in the worklist.

\begin{definition}[Worklist Measures]
\begin{gather*}
\begin{aligned}
|\Gm|_e &= \textstyle\sum_{e\Lto A\in\Gm}|e|_e + \sum_{e\To_a \jg\in\Gm}|e|_e +
    \sum_{\appInfAlg{A}{e}\in\Gm}|e|_e\\
\text{where } |x|_e &= |()|_e = 1 \qquad |\lam e|_e = |e|_e + 1\\
|e_1~e_2|_e &= |e_1|_e + |e_2|_e + 1 \qquad |e:A|_e = |e| + 1\\[2mm]
%
|\Gm|_\Leftrightarrow &= 2\cdot\#_{e\Lto A\in\Gm} +
    \#_{e\To_a \jg\in\Gm} + 3\cdot\#_{\appInfAlg{A}{e}\in\Gm}\\[2mm]
|\Gm|_\forall &= \textstyle\sum_{e\Lto A\in\Gm}|A|_\forall + \sum_{\appInfAlg{A}{e}\in\Gm}|A|_\forall +
    \sum_{A\le B\in\Gm} |A|_\forall + |B|_\forall\\
\text{where } |1|_\forall &= |a|_\forall = |\al|_\forall = 1\\
|A\to B|_\forall &= |A|_\forall + |B|_\forall \qquad |\all A|_\forall = |A|_\forall + 1\\[2mm]
%
|\Gm|_\to &= \textstyle\sum_{e\Lto A\in\Gm}|A|_\to + \sum_{\appInfAlg{A}{e}\in\Gm}|A|_\to +
    \sum_{A\le B\in\Gm} |A|_\to + |B|_\to\\
\text{where } |1|_\to &= |a|_\to = |\al|_\to = 1\\
|A\to B|_\to &= |A|_\to + |B|_\to + 1 \qquad |\all A|_\to = |A|_\to\\[2mm]
%
|\Gm|_{\al} &= \#_{\al\in\Gm}\\
\end{aligned}
\end{gather*}
\end{definition}
\bruno{The definitions look a bit intimidating. I do not have immediate suggestions,
  but perhaps Tom has suggestions on how to improve the presentation here.}

It is not difficult to see that all but two algorithmic reduction rules
decrease the group of measures.
(The result of Rule 29 could be directly reduced by Rule 28, which decreases the measures.)
The two exceptions are Rules 10 and 11. Both rules increase the number of existential variables
without decreasing the number of universal quantifiers.
However, at the same time, they both break the subtyping problem into two sub-problems
(Rule 7 applied right after Rule 10 or 11), in smaller sizes.
Therefore we argue that such split makes the worklist easier, by analyzing a subset of the worklist,
which we call \emph{instantiation judgments} $\Gm_i$ (defined with the help of $\Gm_\le$ for the proof).

\begin{definition}[$\Gm_\le, \Gm_i$]
\begin{gather*}
\begin{aligned}
\Gm_\le &:= \nil \mid \Gm_\le, a \mid \Gm_\le, \al \mid \Gm_\le, A\le B\\
\Gm_i &:= \nil \mid \Gm_i, \al\le A \mid \Gm_i, A\le\al \quad
    \text{where } \al\notin FV(A) \cup FV(\Gm_i)
\end{aligned}
\end{gather*}
\end{definition}

\begin{figure}
\framebox{$\Gm\jExt\Gm'$} $\Gm$ extends to $\Gm'$.
\begin{gather*}
\inferrule*[right=$\mathtt{\jExt id}$]
    {~}{\Gm\jExt\Gm}
\qquad
\inferrule*[right=$\mathtt{\jExt solve}$]
    {|A|_\forall = 0 \quad \Gm_L,[A/\al]\Gm_R \jExt \Gm'}{\Gm_L,\al,\Gm_R \jExt \Gm'}
\qquad
\inferrule*[right=$\mathtt{\jExt \al}$]
    {\Gm_L,\al,\Gm_R \jExt \Gm'}{\Gm_L,\Gm_R \jExt \Gm'}
\end{gather*}
\caption{Worklist Extension}\label{fig:worklist_ext}
\end{figure}

\paragraph{Intuition of instantiation decidability}
These instantiation judgments are the ones consumed and produced by Rules 10 and 11.
When there are universal quantifiers in the initial instantiation judgments $\Gm_i$,
the worklist $\Gm,\Gm_i$ will be multi-reduced to match Rule 8 or 9,
which decreases the number of universal quantifiers by 1, and returns a $\Gm',\Gm_\le$.
Otherwise, no more variables will be introduced,
and each instantiation judgment solves one existential variable,
thus finally there will be $|\Gm_i|$ existential variables solved,
with the worklist reduced to $\Gm'$.
$\Gm'$ basically refers to the original part of worklist $\Gm$.
During the reduction, minor changes to the worklist may occur, such as
existential variable solving or insertion and instantiation.
Nevertheless, the main structure of remains unchanged.

Shown in Figure~\ref{fig:worklist_ext},
worklist extension formally describes the minor changes to worklist during algorithmic reduction.
When the algorithm processes the worklist, the first judgment is processed,
and the rest of the worklist roughly preserves its shape.
Worklist extension is formally defined by two cases:
existential variable solving ($\mathtt{\jExt solve}$) and insertion ($\mathtt{\jExt \al}$).
Existential variable instantiation, as another common form of worklist manipulation,
i.e. $\Gm_L,\al,\Gm_R \jExt \Gm_L, \al[1], \al[2], [\al[1] \to \al[2] / \al]\Gm_R$,
can be derived with the combination of the 3 rules.
Within expectation, some worklist measures are not changed modulo worklist extension.

\newcommand{\equivGm}[1]{|\Gm|_{#1} = |\Gm'|_{#1}}
\begin{lemma}[Measure Invariants of Worklist Extension]
If $\Gm\sto\Gm'$ then $\equivGm{e}$ and $\equivGm\Leftrightarrow$ and $\equivGm\forall$.
\end{lemma}

With all the definitions, we are able to prove the following instantiation decidability lemma.

\begin{lemma}[Instantiation Decidability]\label{lemma:inst:decidable}
For any well-formed algorithmic worklist $(\Gm, \Gm_i)$,
\begin{enumerate}[1)]
    \item When $|\Gm_i|_\forall = 0$, $(\Gm, \Gm_i) \redto \Gm'$,
        where $|\Gm'|_{\al} = |\Gm|_{\al} - |\Gm_i|$ and $\Gm\sto\Gm'$.
    \item When $|\Gm_i|_\forall > 0$, $(\Gm, \Gm_i) \redto (\Gm', \Gm_\le)$,
        where $|\Gm_\le|_\forall = |\Gm_i|_\forall - 1$ and $\Gm\sto\Gm'$.
\end{enumerate}
\end{lemma}

In short, after reducing any instantiation judgment prefix $\Gm_i$,
the algorithm either decreases the number of $\forall$'s
or solves one existential variable per instantiation judgment.
The proof of this lemma is by induction on the measure $2*|\Gm_i|_\to + |\Gm_i|$
of the instantiation judgment list $\Gm_i$.

In summary, the decidability theorem is shown through a lexicographic group of induction measures
$\langle |\Gm|_e, |\Gm|_\Leftrightarrow, |\Gm|_\forall, |\Gm|_{\al}, |\Gm|_\to + |\Gm| \rangle$.
The critical case is that, whenever we encounter an instantiation judgment at the top of the worklist,
we make use of Lemma~\ref{lemma:inst:decidable}, which reduces $|\Gm|_{\al}$ or $|\Gm|_\forall$
by consuming that instantiation judgment. Other cases are relatively straightforward.

Combining all the three main results, we know that the declarative system is decidable,
with the help of our algorithm.
\begin{corollary}[Decidability of Declarative Typing]
Given \emph{wf }$\Om$, it is decidable whether $\Om\redto\nil$ or not.
\end{corollary}
\subsection{Abella}

>>>>

\section{Metatheory}\label{sec:metatheory}

@@

%-------------------------------------------------------------------------------
\subsection{Termination and Decidability}

Finally, we show that our algorithm terminates. In other words:
\begin{theorem}[Decidability]
Given \emph{wf }$\Gm$, it is decidable whether $\Gm\redto\nil$ or not.
\end{theorem}
Our termination proof is based on a lexicographic group of induction measures\\
$\langle |\Gm|_e, |\Gm|_\Leftrightarrow, |\Gm|_\forall, |\Gm|_{\al}, |\Gm|_\to + |\Gm| \rangle$
on the worklist $\Gm$. The definitions of these measures can be found in Figure~\ref{fig:measures}. The first two, 
$|\cdot|_e$ and $|\cdot|_\Leftrightarrow$, measure the total size of terms
and the total difficulty of judgments, respectively. In the latter, check judgements
count for 2, inference judgments for 1 and function inference judgments for 3.
The next two measures, $|\cdot|_\forall$ and $|\cdot|_\to$, count the total number of
universal quantifiers and function types, respectively. Finally,
$|\cdot|_{\al}$ counts the number of existential variables in the worklist.

\begin{figure}[t]
\hfill \framebox{$|\Gm|_e$} \hfill  \llap{Size measure.}
\begin{gather*}
\begin{aligned}
|\Gm|_e &= \textstyle\sum_{e\Lto A\in\Gm}|e|_e + \sum_{e\To_a \jg\in\Gm}|e|_e +
    \sum_{\appInfAlg{A}{e}\in\Gm}|e|_e\\
\text{where } |x|_e &= |()|_e = 1 \qquad |\lam e|_e = |e|_e + 1\\
|e_1~e_2|_e &= |e_1|_e + |e_2|_e + 1 \qquad |e:A|_e = |e| + 1
\end{aligned}
\end{gather*}
\hfill \framebox{$|\Gm|_\Leftrightarrow$} \hfill  \llap{Judgment measure.}
\begin{gather*}
\begin{aligned}
|\Gm|_\Leftrightarrow &= 2\cdot\#_{e\Lto A\in\Gm} +
    \#_{e\To_a \jg\in\Gm} + 3\cdot\#_{\appInfAlg{A}{e}\in\Gm}
\end{aligned}
\end{gather*}
\hfill \framebox{$|\Gm|_\forall$} \hfill  \llap{Polymorphism measure.}
\begin{gather*}
\begin{aligned}
|\Gm|_\forall &= \textstyle\sum_{e\Lto A\in\Gm}|A|_\forall + \sum_{\appInfAlg{A}{e}\in\Gm}|A|_\forall +
    \sum_{A\le B\in\Gm} |A|_\forall + |B|_\forall\\
\text{where } |1|_\forall &= |a|_\forall = |\al|_\forall = 1\\
|A\to B|_\forall &= |A|_\forall + |B|_\forall \qquad |\all A|_\forall = |A|_\forall + 1
\end{aligned}
\end{gather*}
\hfill \framebox{$|\Gm|_\to$} \hfill  \llap{Function measure.}
\begin{gather*}
\begin{aligned}
|\Gm|_\to &= \textstyle\sum_{e\Lto A\in\Gm}|A|_\to + \sum_{\appInfAlg{A}{e}\in\Gm}|A|_\to +
    \sum_{A\le B\in\Gm} |A|_\to + |B|_\to\\
\text{where } |1|_\to &= |a|_\to = |\al|_\to = 1\\
|A\to B|_\to &= |A|_\to + |B|_\to + 1 \qquad |\all A|_\to = |A|_\to
\end{aligned}
\end{gather*}
\hfill \framebox{$|\Gm|_{\al}$} \hfill  \llap{Existential measure.}
\begin{gather*}
\begin{aligned}
|\Gm|_{\al} &= \#_{\al\in\Gm}\\
\end{aligned}
\end{gather*}
\caption{Worklist measures}
\label{fig:measures}
\end{figure}

It is not difficult to see that all but two algorithmic reduction rules
decrease the group of measures.  (The result of Rule 29 could be directly
reduced by Rule 28, which decreases the measures.) The two exceptions are Rules
10 and 11. Both rules increase the number of existential variables without
decreasing the number of universal quantifiers. However, they are both
immediately followed by Rule 7, which breaks the subtyping problem into two
smaller problems of the form $\al \le A$ and $A \le \al$ which we call
\emph{instantiation judgements}.

We now show that entirely reducing these smaller problems leaves the worklist
in a state with an overall smaller measure. Our starting point is a 
worklist of the form $\Gm,\Gm_i$ where $\Gm_i$ are instantiation judgments.
\begin{gather*}
\begin{aligned}
\Gm_i &:= \nil \mid \Gm_i, \al\le A \mid \Gm_i, A\le\al \quad
    \text{where } \al\notin FV(A) \cup FV(\Gm_i)
\end{aligned}
\end{gather*}
Fully reducing these instantiation judgments at the top of the worklist
has a twofold impact. Firstly, new entries may be pushed onto the worklist which
are not instantiation judgments. This only happens when $\Gm_i$ contains a universal quantifier
that is reduced by Rule 8 or 9. The new entries then are of the form $\Gm_\forall$:
\begin{gather*}
\begin{aligned}
\Gm_\le &:= \nil \mid \Gm_\le, a \mid \Gm_\le, \al \mid \Gm_\le, A\le B
\end{aligned}
\end{gather*}
Secondly, reducing the instantiation judgments may also affect the remainder of the worklist $\Gm$,
by solving existing existentials and introducing new ones. This worklist update is captured in the 
update judgement $\Gm \jExt \Gm'$ defined in Figure~\ref{fig:worklist_ext}.
For instance, existential variable instantiation, 
i.e. $\Gm_L,\al,\Gm_R \jExt \Gm_L, \al[1], \al[2], [\al[1] \to \al[2] / \al]\Gm_R$,
can be derived as a combination of the three rules that define the update relation.

\begin{figure}
\hfill \framebox{$\Gm\jExt\Gm'$} \hfill \llap{$\Gm$ updates to $\Gm'$}.
\begin{gather*}
\inferrule*[right=$\mathtt{\jExt id}$]
    {~}{\Gm\jExt\Gm}
\qquad
\inferrule*[right=$\mathtt{\jExt solve}$]
    {|A|_\forall = 0 \quad \Gm_L,[A/\al]\Gm_R \jExt \Gm'}{\Gm_L,\al,\Gm_R \jExt \Gm'}
\qquad
\inferrule*[right=$\mathtt{\jExt \al}$]
    {\Gm_L,\al,\Gm_R \jExt \Gm'}{\Gm_L,\Gm_R \jExt \Gm'}
\end{gather*}
\caption{Worklist Update}\label{fig:worklist_ext}
\end{figure}

The good news is that worklist update does not affect the three main worklist measures:
\newcommand{\equivGm}[1]{|\Gm|_{#1} = |\Gm'|_{#1}}\label{lemma:inst:invariant}
\begin{lemma}[Measure Invariants of Worklist Extension]
If $\Gm\sto\Gm'$ then $\equivGm{e}$ and $\equivGm\Leftrightarrow$ and $\equivGm\forall$.
\end{lemma}

Moreover, we can characterise the reduction of the instantiation judgments as follows.
\begin{lemma}[Instantiation Decidability]\label{lemma:inst:decidable}
For any well-formed algorithmic worklist $(\Gm, \Gm_i)$,
\begin{enumerate}[1)]
    \item When $|\Gm_i|_\forall = 0$, $(\Gm, \Gm_i) \redto \Gm'$,
        where $|\Gm'|_{\al} = |\Gm|_{\al} - |\Gm_i|$ and $\Gm\sto\Gm'$.
    \item When $|\Gm_i|_\forall > 0$, $(\Gm, \Gm_i) \redto (\Gm', \Gm_\le)$,
        where $|\Gm_\le|_\forall = |\Gm_i|_\forall - 1$ and $\Gm\sto\Gm'$.
\end{enumerate}
\end{lemma}
Hence, reducing the instantiation judgment prefix $\Gm_i$ either
decreases the number of universal quantifiers or 
the algorithm either decreases the number of $\forall$'s
or solves one existential variable per instantiation judgment.
The proof of this lemma proceeds by induction on the measure $2*|\Gm_i|_\to + |\Gm_i|$
of the instantiation judgment list $\Gm_i$.

Let us go back to the whole algorithm and summarize our findings.
The decidability theorem is shown through a lexicographic group of induction measures
$\langle |\Gm|_e, |\Gm|_\Leftrightarrow, |\Gm|_\forall, |\Gm|_{\al}, |\Gm|_\to + |\Gm| \rangle$
which is decreased by nearly all rules.
In the exceptional case that the measure does not decrease immediately, 
we encounter an instantiation judgment at the top of the worklist. We can then
make use of Lemma~\ref{lemma:inst:decidable} to show that $|\Gm|_{\al}$ or $|\Gm|_\forall$ decreases
when that instantiation judgment is consumed. Moreover, Lemma~\ref{lemma:inst:invariant} establishes
that no higher-priority measure component increases. Hence, in the exceptional case we have an overall measure decrease too.

Combining all three main results (soundness, completeness and termination), we conclude that the declarative system is decidable
by means of our algorithm.
\begin{corollary}[Decidability of Declarative Typing]
Given \emph{wf }$\Om$, it is decidable whether $\Om\redto\nil$ or not.
\end{corollary}
\subsection{Abella}
